% Genomes Workflow Module Template - LBCM UFRGS
% Based on NotesTeX and Gilles Castel's aesthetic approach
% Module: 02 - Data pre-processing
% Date: DATE_PLACEHOLDER

\documentclass[a4paper,11pt]{article}

% Use ONLY the beautiful notes preamble for notes
\input{preamble.tex}

\title{
    \vfill
    \textcolor{noteblue}{\Huge Genomes Workflow - LBCM}\\
    \vspace{5.0cm}
    \textcolor{sidenotegray}{\huge 02 - Data pre-processing}\\
    \vspace{1.5cm}
    \textcolor{sidenotegray}{\LARGE }\\
    \vspace{5.0cm}
    \textcolor{sidenotegray}{\large \today}
    \vfill
}
\author{}
\date{}

\begin{document}

\maketitle
\thispagestyle{empty}
\newpage
\tableofcontents
\newpage
\section{Overview}
Once the base introduction and general guidelines were covered, we can start
working directly on the pipeline itself. For starters, it's always advised to
pre-process our data, that being: understanding it's origin, how we expect it to
behave and what special treatment does it need. Following such general stone
path can help we obtain better results later on, securing data integrity and
usability.
\subsection{Module Objectives}
\begin{itemize}
    \item Present the logic behind data pre-processing.
    \item Approach tools that operate on such function.
    \item Analyze output examples.
    \item Comprehend the importance of such step on the general workflow.
\end{itemize}
\subsection{Module Required Tools}
\begin{itemize}
    \item Conda (recommended)
    \item Git (recommended)
    \item FASTP
    \item FASTQC
\end{itemize}
\subsection{Module Structure}
The current module is divided in two big sections:
\begin{description}
    \item[Theoretical Framework:] Approaches the theory behind the process
        that's being mainly covered on the current module. The idea is to allow
        the user to consult for general concepts comprehension and, if needed,
        guide him to sources for a more in depth coverage.
    \item [Practical approach:] In this section, we shall explain in practice
        how this step can be applied. For such, NCYC357.fastq, a paired-end
        Illumina set of reads, will be used, although the intrinsic logic and
        observations can be applied to other samples.
\end{description}
\section{Theoretical Framework - Data pre-processing}
\subsection{What is it?}
\subsection{Why it's important?}
\subsection{Base assumptions}
\section{Practical approach}
To better understand the steps that, combined, allows us to comprehend and
optimize our data, we're gonna apply the general pipeline
\subsection{Data analysis}
\subsubsection{Gathering the data}
\subsubsection{Understanding our data}
\subsubsection{Taking decisions}
\subsection{Data pre-processing}
\subsubsection{Choosing the right tools}
\subsubsection{How results can become parameters}
\subsubsection{Pre-processing}
\subsubsection{Understanding what's found}


% BIBLIOGRAPHY
\newpage
\printbibliography

\end{document}
