% Genomes Workflow Module Template - LBCM UFRGS
% Based on NotesTeX and Gilles Castel's aesthetic approach
% Module: 01 - Introduction
% Date: DATE_PLACEHOLDER

\documentclass[a4paper,11pt]{article}

% Use ONLY the beautiful notes preamble for notes
\input{preamble.tex}

\title{
    \vfill
    \textcolor{noteblue}{\Huge Genomes Workflow - LBCM}\\
    \vspace{5.0cm}
    \textcolor{sidenotegray}{\huge 01 - Introduction}\\
    \vspace{1.5cm}
    \textcolor{sidenotegray}{\LARGE General guidelines}\\
    \vspace{5.0cm}
    \textcolor{sidenotegray}{\large \today}
    \vfill
}
\author{}
\date{}

\begin{document}

\maketitle
\thispagestyle{empty}
\newpage

\section{Overview}

\subsection{Module Objectives}
\sn{Learning goals}

This module covers:
\begin{itemize}
    \item Repository organization.
    \item Available scripts.
    \item General pipeline structure.
    \item Modules organization.
\end{itemize}

\section{Background}

\subsection{General Context}
\sn{Why this matters}

Key concepts:
\begin{itemize}
    \item The integration between biological knowledge and IT or
        bioinformatics tools can be harsh.
    \item The presence and interference of multiple key concepts of each area
        creates a big wall on newcomers.
    \item Explain key features and provide support in a organized way can help
        to circumvent this initial barrier.
\end{itemize}

\section{Workflow \& Methods}

\subsection{The repository}
\sn{Key parameters}
The repository is structured aiming to facilitate user operation and base codes
and modules maintenance. The main separation happens in the global folder, that
stores templates, bibliography files and scripts for project management.\par
The base skeleton of the repository can be seen below:\par
\begin{verbatim}
.
|-- README.md
|-- global
|   |-- bibliography
|   |   `-- master.bib
|   |-- scripts
|   |   |-- main.py
|   |   `-- tools
|   |       |-- __init__.py
|   |       |-- create_module.py
|   |       `-- set_vim_config.py
|   `-- templates
|       |-- README_template.md
|       |-- enhanced_vimtex_config.vim
|       |-- gilles_snippets.snippets
|       |-- main_template.tex
|       `-- preamble_template.tex
|-- modules
|   `-- 01_introduction
|       |-- README.md
|       |-- images
|       |-- main.tex
|       |-- preamble.tex
|       `-- references.bib
|-- requirements.txt
|-- scripts
|   |-- README.md
|   |-- build_course.py
|   |-- config.yaml
|   |-- interactive_builder.py
|   |-- main.py
|   |-- requirements.txt
|   |-- tools
|   |   `-- __init__.py
|   `-- utils
|       |-- __init__.py
|       |-- file_manager.py
|       |-- latex_compiler.py
|       |-- module_validator.py
|       `-- pdf_merger.py
`-- setup.py
\end{verbatim}
\subsection{Available scripts}
Currently, there are two distinct scripts folders. The first, located inside the
global folder, contains py scripts focused on project maintenance, while the
second receives all py scripts user-oriented.
\subsubsection{Global - Scripts}
\begin{description}
    \item[main.py] Presents a selection menu to all available scripts inside the
        tools folder.
    \item[create\_module.py] Allows the automatic creation of a new Module
        folder. Includes the basic structure, with a README file, main and
        preamble tex files, images folder and a local references.bib file.
    \item[set\_vim\_config.py] Change the user .vimrc file to a custom pre-made
        one, that can be found inside the templates folder. Allows VimTeX
        and UltiSnips integration, alongside keyboard shortcuts for easier
        navigation.
\end{description}
\subsubsection{Main folder - Scripts}
\begin{example}
Practical example with real genomic data.
\end{example}

\section{Practical Examples}

\subsection{Example 1: Basic Analysis}
\sn{Input/output files}

\begin{itemize}
    \item \textbf{Input:} Sample data description
    \item \textbf{Command:} Based on approach from \cite{example2024}
    \item \textbf{Output:} Expected results and file formats
\end{itemize}

\begin{lstlisting}[language=bash, caption=Basic command example]
# Example command with typical genomic data
tool_name -i input_file.fasta -o output_file.txt --parameter value
\end{lstlisting}

\subsection{Example 2: Advanced Usage}
\sn{Complex parameters}

\begin{lstlisting}[language=bash, caption=Advanced analysis pipeline]
# Multi-step analysis pipeline
step1_tool input.fasta | step2_tool --param1 value1 > intermediate.txt
step3_tool intermediate.txt --param2 value2 -o final_result.txt
\end{lstlisting}

\section{Results \& Interpretation}

\subsection{Output Files}
\sn{File formats}

Common output formats and their interpretation:
\begin{itemize}
    \item \textbf{Format 1:} Description and typical contents
    \item \textbf{Format 2:} When and how to use this output
    \item \textbf{Quality metrics:} How to assess result quality
\end{itemize}

\begin{remark}
Important note about result interpretation following \citet{author2024}.
\end{remark}

\section{Scripts \& Code}

\subsection{Helper Scripts}
\begin{lstlisting}[language=Python, caption=Data processing script]
#!/usr/bin/env python3
"""
Helper script for genomic data processing
Usage: python script.py input.fasta output.txt
"""

def process_sequences(input_file, output_file):
    """Process genomic sequences"""
    with open(input_file, 'r') as f:
        sequences = f.read()
    
    # Processing logic here
    processed = sequences.upper()
    
    with open(output_file, 'w') as f:
        f.write(processed)

if __name__ == "__main__":
    import sys
    process_sequences(sys.argv[1], sys.argv[2])
\end{lstlisting}

\subsection{Quality Control}
\begin{lstlisting}[language=bash, caption=QC pipeline]
#!/bin/bash
# Quality control pipeline for genomic data

# Check file format
file_format_check.py $INPUT_FILE

# Basic statistics
sequence_stats.py $INPUT_FILE > stats.txt

# Quality assessment
quality_assessment_tool $INPUT_FILE --output qc_report.html
\end{lstlisting}

\section{Troubleshooting \& Best Practices}

\subsection{Common Issues}
\sn{Error solutions}

\begin{itemize}
    \item \textbf{Memory errors:} Reduce dataset size or increase available RAM
    \item \textbf{Format issues:} Check input file formatting and encoding
    \item \textbf{Parameter tuning:} Guidelines for optimization
\end{itemize}

\subsection{Best Practices}
\begin{itemize}
    \item \textbf{Data backup:} Always keep original data copies
    \item \textbf{Version control:} Track analysis versions and parameters
    \item \textbf{Documentation:} Record all analysis steps and decisions
    \item \textbf{Reproducibility:} Use consistent environments and seeds
\end{itemize}

\begin{references}
    \item Key papers: \cite{example2024,author2024,smith2024}
    \item Software documentation: [Tool official docs]
    \item Related modules: [Other workflow modules]
\end{references}

\section{Exercises \& Next Steps}

\begin{itemize}
    \item \todo{Practice with provided sample data}
    \item \todo{Try different parameter settings}
    \item \important{Apply to your own genomic dataset}
    \item \todo{Explore advanced features}
\end{itemize}

\begin{notes}
Additional observations and module-specific notes...

\highlight{Key insight}: Connection between this tool and genome annotation pipeline

\idea{Extension}: Integration with other bioinformatics tools in the workflow

\end{notes}

% BIBLIOGRAPHY
\newpage
\printbibliography

\end{document}
